Non-word repetition in Yélî Dnye

### 2020-12-18 manuscript.tex
minor typo fixes

\abstract{
In nonword repetition (NWR) studies, participants are presented
auditorily with an item that is phonologically legal but lexically
meaningless in the local language. NWR scores are thought to reflect
long-term phonological knowledge as well as online phonological working
memory and flexible production patterns. In this study, we report on NWR
results among children learning Yélî Dnye, an isolate spoken on Rossel
Island in Papua New Guinea, with an unusually dense phonological
inventory. This study contributes to three lines of research. First, we
document that non-word items containing typologically rare sounds are
repeated without changes less often that non-words containing more
common sounds. Second, we document rather weak effects of item length,
contributing to other research suggesting that length effects may be
language-specific. Third, we do not find strong individual variation
effects in this population, contrary to previous results documenting
strong age-related effects. Together, these data provide a unique view
of online phonological processing in a seldom-studied language, and
contribute to both typological and language acquisition research.

@@

\begin{itemize}
\tightlist
\item
  double check my frequency entries in segments.xlsx using the search
  function in \url{https://phoible.org/parameters}; take a look at
  \url{http://phoible.github.io/conventions/} in case you see something
  about double articulation (I didn't find tp or lBj) 
\item
  draft discussion
\end{itemize}

@@

NWR has been used to seek answers to a variety of theoretical questions,
including what the links are between phonology, working memory, and the
lexicon (Bowey, 2001), and how extensively apparent phonological
constraints found in the lexicon affect online production (Gallagher,
2014). NWR is also frequently used in applied contexts, notably as a
diagnostic tool for language delays and disorders (Estes, Evans, \&
Else-Quest, 2007). Since non-words can be generated in any language, it
has attracted the attention of researchers working in multilingual and
linguistically diverse environments, particularly in Europe (Action,
2009; Meir, Walters, \& Armon-Lotem, 2016). However, it has seldom been
used outside of Europe and North America (with exceptions including
Gallagher, 2014, and Cristia, Farabolini, Scaff, Havron, and Stieglitz
(2020)). Further, the languages it has been used with have historically
had small to moderate phonological inventories (Maddieson, 2005).

In the present study, we use NWR to investigate the phonological
development of children learning Yélî Dnye, an isolate language spoken
in Papua New Guinea (PNG) that has a large and unusually dense
phonological inventory. The study was designed to contribute to multiple
aspects of our understanding of phonological development: First, we
included a subset of non-word items with typologically rare and/or
challenging sounds to ask whether these rare sounds are disadvantaged in
perception and/or production, both in terms of overall repetition scores
and patterns of mispronunciation. Second, we varied the length (in
syllables) of non-words to investigate the impact of word length on
repetition and mispronunciation; previous NWR research has shown
variable effects of length between populations. Third, these data give a
first impression of how this task is responded to in a rarely-studied
language and culture; we add to this by further investigating whether
there are structured sources that account for individual variation in
our sample.

@@

Most words in Yélî Dnye are bisyllabic (\textasciitilde{}50\%), with
monosyllabic words (\textasciitilde{}40\%) appearing most commonly after
that, and with tri-and-above syllabic words appearing least frequently
(\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the most
recent Yélî Dnye dictionary at time of writing, Levinson, accepted). The
vast majority of syllables use a CV format. A small portion of the
lexicon features words with a final closed syllable, limited to codas of
-/m/, -/p/, or -/j/ (e.g., `ndap' (Spondylus shell) /ṇṭæp/). However, in
spontaneous speech an epenthetic /ɯ/ is often inserted after word-final
coda, yielding a CVCV structure instead (e.g., `ndapî' /`ṇṭæ.pɯ/). This
process is used frequently with English loan words that have a coda
(e.g., 'ponî' (phone) /`pɔ.nɯ/). An even smaller portion of the lexicon
features words starting with a vowel (e.g., 'ala' (here) /æ.'læ/), but
these are limited to /æ/-. Finally, the lexicon features a handful of
single-vowel grammatical morphemes (see Levinson (accepted) for
details).

@@

Laying 250 nautical miles off the coast of mainland PNG and surrounded
by a barrier reef, transport to and from the island is both infrequent
and irregular. There is no electricity system (we use solar panels) or
motor vehicles on the island, save dinghys with outboard motors that,
even when there is deisel available, are primarily reserved for medical
transport. There is a medical radio connection at the health clinic,
located at the Catholic Mission, via which news and messages are often
passed to and from mainland PNG. There is also a cellular tower on the
island, though it is often inoperative, usually for multiple months or
years at a time. As such, access to news and outside connections,
including international phone calls and digital exchanges that require
significant data transfer, is typically unavailable. Our data collection
is therefore limited to the duration of the researchers' on-island
visits.

Despite restricted outside contact, formal education is a priority for
many Yélî families. A recent study surveying more than 40 parents of
young children in this region of the island found that nearly all
parents completed 6 or more years of education, with 65\% of mothers and
49\% of fathers completing the full range of locally available education
(\textasciitilde{}9 years), with around half of those parents having
pursued an additional 2+ years of education on other islands in the
region or on mainland PNG (Casillas, Brown, \& Levinson, 2020). While
there are a handful of local schools around the island, these may be
well out of walking distance for many children (i.e., more than 1 hour
on foot or by canoe each day). It is very common for households situated
close to schools to billet their school-aged relatives during the
weekdays (sometimes the weekends too, if the child's home is very far)
for long segements of each school year. Combined with practices of
collective child guardianship within each close-knit hamlet, the
practical consequence of this billeting for researchers is that adult
consent can often come from a combination of aunts, uncles, adult
cousins, and grandparents standing in for the child's biological
parents. On top of this, child assent is culturally pertinent, as child
independence is encouraged and respected from toddlerhood onward (Brown
\& Casillas, in press).\footnote{For these reasons, we were unable to
  record precise details for a few children, including child date of
  birth, parent age in years, and parent education.} Finally, as
mentioned, the language of schooling in this region is English, and so
the majority of teens and adults on the island speak some English---many
speak it very fluently.

@@

In a basic NWR task, the participant listens to a production of a
word-like form, such as /bilik/, and then repeats back what they heard
without changing any phonological feature that is contrastive in the
language. For instance, in English, a response of {[}bilig{]} or
{[}pilik{]} would be scored as incorrect; a response {[}biilik{]}, where
the vowel is lengthened without change of quality would be scored as
correct, because English does not have a general feature of contrastive
vowel length. There is some variation in how past NWR studies have
designed the presentation procedure and structure of items. For example,
while items are often presented orally by the experimenter (Torrington
Eaton, Newman, Ratner, \& Rowe, 2015), an increasing number of studies
have turned instead to playing back pre-recorded stimuli in order to
increase control in stimulus presentation (Brandeker \& Thordardottir,
2015). Additionally, while some studies have used 10-15 non-words,
others have employed up to 46 unique items (Piazzalunga, Previtali,
Pozzoli, Scarponi, \& Schindler, 2019). Authors also often modulate
structural complexity, typically measured in terms of item length
(measured in number of syllables) and/or syllable structure (open as
opposed to closed syllables, Gallon, Harris, \& Van der Lely, 2007).

Previous work typically steers clear of articulartorily and/or
acoustically challenging sounds, but we included some in our experiment
to more adequately represent Yélî Dnye's phonology. We ultimately used a
relatively large number of items that would enable us to explor both
variation in structural complexity and in more vs.~less challenging
sounds. However, aware that this large item inventory might render the
task longer and more tiresome, we split items across children (see
below). Naturally, designing the task in this way may render the study
of individual variation within the population more difficult because
different children are exposed to different items. However, a review of
previous work on individual variation suggested to us that effects of
individual differences in NWR are relatively small, and would not be
detectable with the sample size that we could collect during our short
visit. That said, we contribute to the literature by also reporting
descriptive analyses of individual variation that could potentially be
integrated in meta-analytic efforts.

@@

\begin{itemize}
\tightlist
\item
  Does the cross-linguistic frequency of sounds predict NWR scores? Are
  rarer sounds more often substituted by commoner sounds?
\item
  How does score change as a function of item length in number of
  syllables?
\item
  Is individual variation attributable to child age, sex, birth order,
  monolingual status, and/or parental education?
\end{itemize}

>>>>

\abstract{
In nonword repetition (NWR) studies, participants are presented
auditorily with an item that is phonologically legal but lexically
meaningless in their language. NWR scores are thought to reflect
long-term phonological knowledge as well as online phonological working
memory and flexible production patterns. In this study, we report on NWR
results among children learning Yélî Dnye, an isolate spoken on Rossel
Island in Papua New Guinea that has an unusually dense phonological
inventory. This study contributes to three lines of research. First, we
document that non-word items containing typologically frequent sounds
are repeated without changes more often that non-words containing
typologically rare sounds. Second, we document rather weak effects of
item length, contributing to other research suggesting that length
effects may be language-specific. Third, we do not find strong
individual variation effects in this population, contrary to previous
results documenting strong age-related effects. Together, these data
provide a unique view of online phonological processing in an
understudied language while making preliminary connections between
patterns in language development and in the frequency of linguistic
features in human language at large.

@@

\begin{itemize}
\tightlist
\item
  draft discussion
\end{itemize}

@@

NWR has been used to seek answers to a variety of theoretical questions,
including what the links are between phonology, working memory, and the
lexicon (Bowey, 2001), and how extensively phonological constraints
found in the lexicon affect online production (Gallagher, 2014). NWR is
also frequently used in applied contexts, notably as a diagnostic tool
for language delays and disorders (Estes, Evans, \& Else-Quest, 2007).
Since non-words can be generated in any language, it has attracted the
attention of researchers working in multilingual and linguistically
diverse environments, particularly in Europe (Action, 2009; Meir,
Walters, \& Armon-Lotem, 2016). However, it has seldom been used outside
of Europe and North America (with exceptions including Gallagher, 2014,
and Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020)). Further,
the languages it has been used with have historically had small to
moderate phonological inventories (Maddieson, 2005).

In the present study, we use NWR to investigate the phonological
development of children learning Yélî Dnye, an isolate language spoken
in Papua New Guinea (PNG) that has a large and unusually dense
phonological inventory. The study was designed to contribute to multiple
aspects of our understanding of phonological development: First, we
included a subset of non-word items with typologically rare and/or
challenging sounds to ask whether these rare sounds are disadvantaged in
perception and/or production, both in terms of overall repetition scores
and patterns of mispronunciation. Second, we varied the length (in
syllables) of non-words to investigate the impact of word length on
repetition and mispronunciation; previous NWR research has shown
variable effects of length between populations. Third, these data give a
first impression of how this task is responded to in a rarely-studied
language and cultural setting; we add to this contribution by further
investigating whether there are structured sources that account for
individual variation in our sample.

@@

Most words in Yélî Dnye are bisyllabic (\textasciitilde{}50\%), with
monosyllabic words (\textasciitilde{}40\%) appearing most commonly after
that, and with tri-and-above syllabic words appearing least frequently
(\textasciitilde{}10\%; based on \textgreater{}5800 lexemes in the most
recent Yélî Dnye dictionary at time of writing; Levinson, accepted). The
vast majority of syllables use a CV format. A small portion of the
lexicon features words with a final closed syllable, limited to codas of
-/m/, -/p/, or -/j/ (e.g., `ndap' (Spondylus shell) /ṇṭæp/). However, in
spontaneous speech an epenthetic /ɯ/ is often inserted after word-final
coda, yielding a CV.CV structure instead (e.g., `ndapî' /ṇṭæ.pɯ/). This
process is used frequently with English loan words that have a coda
(e.g., `ponî' (phone) /`pɔ.nɯ/). An even smaller portion of the lexicon
features words starting with a vowel (e.g., 'ala' (here) /æ.læ/), but
these are limited to /æ/-. Finally, the lexicon features a handful of
single-vowel grammatical morphemes (see Levinson (accepted) for
details).

@@

Laying 250 nautical miles off the coast of mainland PNG and surrounded
by a barrier reef, transport to and from the island is both infrequent
and irregular. There is no electricity system (we use solar panels) or
motor vehicles on the island, save dinghys with outboard motors that are
primarily reserved for medical transport when there is diesel available.
There is a medical radio connection at the health clinic, located at the
Catholic Mission, via which news and messages are often passed to and
from mainland PNG. There is also a cellular tower on the island, though
it is often inoperative, usually for multiple months or years at a time.
As such, access to news and outside connections, including international
phone calls and digital exchanges that require significant data
transfer, is typically unavailable. Our data collection is therefore
limited to the duration of the researchers' on-island visits.

Despite restricted outside contact, formal education is a priority for
many Yélî families. A recent study surveying more than 40 parents of
young children in this region of the island found that nearly all
parents completed 6 or more years of education, with 65\% of mothers and
49\% of fathers completing the full range of locally available education
(\textasciitilde{}9 years), with around half of those parents having
pursued an additional 2+ years of education on other islands in the
region or on mainland PNG (Casillas, Brown, \& Levinson, 2020). While
there are a handful of local schools around the island, these may be
well out of walking distance for many children (i.e., more than 1 hour
on foot or by canoe each day). It is very common for households that are
situated close to a school to billet their school-aged relatives during
the weekdays (sometimes the weekends too, if the child's home is very
far) for long segements of each school year. Combined with practices of
collective child guardianship within each close-knit hamlet, the
practical consequence of this billeting for researchers is that adult
consent can often come from a combination of aunts, uncles, adult
cousins, and grandparents standing in for the child's biological
parents. On top of this, child assent is culturally pertinent, as child
independence is encouraged and respected from toddlerhood onward (Brown
\& Casillas, in press).\footnote{For these reasons, we were unable to
  record precise details for a few children, including child date of
  birth, parent age in years, and parent education.} Finally, as
mentioned, the language of schooling in this region is English, and so
the majority of teens and adults on the island speak some English---many
speak it very fluently.

@@

In a basic NWR task, the participant listens to a production of a
word-like form, such as /bilik/, and then repeats back what they heard
without changing any phonological feature that is contrastive in the
language. For instance, in English, a response of {[}bilig{]} or
{[}pilik{]} would be scored as incorrect; a response {[}biːlik{]}, where
the vowel is lengthened without change of quality would be scored as
correct, because English does not have a general feature of contrastive
vowel length. There is some variation in how past NWR studies have
designed the presentation procedure and structure of items. For example,
while items are often presented orally by the experimenter (Torrington
Eaton, Newman, Ratner, \& Rowe, 2015), an increasing number of studies
have turned instead to playing back pre-recorded stimuli in order to
increase control in stimulus presentation (Brandeker \& Thordardottir,
2015). Additionally, while some studies have used 10-15 non-words,
others have employed up to 46 unique items (Piazzalunga, Previtali,
Pozzoli, Scarponi, \& Schindler, 2019). Authors also often modulate
structural complexity, typically measured in terms of item length
(measured in number of syllables) and/or syllable structure (open as
opposed to closed syllables, Gallon, Harris, \& Van der Lely, 2007).

Previous work typically steers clear of articulartorily and/or
acoustically challenging sounds, but we included some in our experiment
to more adequately represent Yélî Dnye's phonology. We ultimately used a
relatively large number of items that would enable us to explore both
variation in structural complexity and in more vs.~less challenging
sounds. However, aware that this large item inventory might render the
task longer and more tiresome, we split items across children (see
below). Naturally, designing the task in this way may make the study of
individual variation within the population more difficult because
different children are exposed to different items. However, a review of
previous work on individual variation suggested to us that effects of
individual differences in NWR are relatively small, and would not be
detectable with the sample size that we could collect during our short
visit. That said, we contribute to the literature by also reporting
descriptive analyses of individual variation that could potentially be
integrated in meta-analytic efforts.

@@

\begin{itemize}
\tightlist
\item
  Does the cross-linguistic frequency of sounds in the stimuli predict
  NWR scores? Are rarer sounds more often substituted by commoner
  sounds?
\item
  How does score change as a function of item length in number of
  syllables?
\item
  Is individual variation in NWR scores attributable to child age, sex,
  birth order, monolingual status, and/or parental education?
\end{itemize}
